Resource Adequacy and Operational Security Interaction in the EPOC 2030-50 Project

### 2022-11-30 papers/main.tex
Made residual load plot, fixing full year ED

\begin{abstract}
    This report outlines an attempt to integrate a resource adequacy and operational security exercise. The idea is to use a unit commitment schedule which is then used as the input to a more detailed power flow model. An additional goal was to make explicit the trade-off between day ahead adequacy and real time reliability. We used an adapted version of KU Leuven's \ac{DUCPR} model which accounts for network constraints in the reserve activation phase for this purpose. However the \ac{DUCPR} model is not able to capture enough detail in the reserve activation phase for this tradeoff to be made. This and other omissions of the \ac{DUCPR} model when compared to the power flow model meant the results of the two models differed. All code for the adequacy side of this exercise can be found at \href{https://github.com/junglegobs/ASoSEPOC}{https://github.com/junglegobs/ASoSEPOC}.
\end{abstract}

@@

Adequacy assessments and operational security analyses are two fundamental exercises in determining and ensuring the reliable operation of the electric power system. The former assesses the ability of the power system to satisfy demand for electricity in the day ahead stage, while the latter assesses (or ensures, depending on the context) that an unexpected event, such as an outage of a transmission line or generator or lower or greater than expected realisation of demand or renewable generation, does not lead to cascading failures and a possible blackout.

Typically, these exercises are performed independently. This interaction combines the two by feeding the results of an adequacy assessment to an operational security analysis.

@@

Secondly, the adequacy assessment is done using an unconventional \ac{UC} model, a modified version of the \ac{DUCPR} model described in~\cite{Bruninx2017}. This model can be seen as a compromise between a computationally expensive stochastic \ac{UC} model and a more tractable though less accurate deterministic \ac{UC} model with operating reserve requirements. The treatment of operating reserves in this model allows a trade-off between shedding load in the day ahead stage, which would reduce day ahead adequacy, and shedding operating reserves, which should reduce operational security. By making this trade-off explicit and verifying the results in an operational security analysis we aim to elaborate on the results of~\cite{Hermans2018}, which did not consider the possibility that shedding operating reserves would lead to insecure operation of the electric power system.

@@

The rest of this document is organised as follows. \secref{sec:model} describes the \ac{DUCPR} model used to simulate power system operations and obtain day ahead adequacy and very approximate real time operational security indicators, namely scheduled load shedding and reserve shedding. \secref{sec:data} describes the stylised Belgian system analysed. \secref{sec:initial_results} analyses a single day in order to identify sources of adequacy and security issues, while \secref{sec:adequacy_vs_security} analyses the trade-off between these two.

@@

Sets, variables and costs related to unit commitment and storage operation have been omitted for brevity though they are included in the model. The interested reader is referred to \href{https://github.com/junglegobs/ASoSEPOC}{https://github.com/junglegobs/ASoSEPOC} for the relevant code.

@@

\begin{itemize}
    \item $G$ - Generators
    \item $G_n$ - Generators located at node $n$
    \item $GD$ - Dispatchable / thermal / conventional generators
    \item $GR$ - Renewable / variable generators (\ac{VRES}).
    \item $S$ - Storage technologies
    \item $N$ - Nodes or buses in the network
    \item $B$ - Lines or branches
    \item $T$ - Time steps / intervals / slices
    \item $L^+$ - Upward reserve levels
    \item $L^-$ - Downward reserve levels
\end{itemize}

@@

\begin{itemize}
    \item $D_{nt}$ - Demand
    \item $PTDF_{nl}$ - Power Transfer Distribution Factor
    \item $AF_{gt}$ - Availability Factor
    \item $K_g$ - Capacity
    \item $P^{min}_g$ - Minimum power output
    \item $P^{max}_g$ - Maximum power output
    \item $MUT_g$ - Minimum up time
    \item $MDT_g$ - Minimum down time
    \item $RAMP_g$ - Maximum ramping limit (up and down)
    \item $D^{L^+}_{lt}$ - Upward reserve requirement
    \item $D^{L^-}_{lt}$ - Downward reserve requirement
    \item $P^{L^+}_{lt}$ - Upward reserve level activation probability
    \item $P^{L^-}_{lt}$ - Downward reserve level activation probability
\end{itemize}

@@

\begin{itemize}
    \item $z_{gt}$ - Commitment
    \item $v_{gt}$ - start up
    \item $w_{gt}$ - Shut down
    \item $q_{gt}$ - Generation
    \item $\hat{q}_{gt}$ - Generation above the minimum stable operating point (0 in the case of renewables).
    \item $d_{st}$ - Storage discharge
    \item $c_{st}$ - Storage charge
    \item $y_{st}$ - Binary variable, 1 if storage discharges and 0 if it charges
    \item $e_{st}$ - Storage state of charge variable
    \item $ls_{nt}$ - Load shedding
    \item $inj_{nt}$ - Node injection
    \item $f_{lt}$ - Branch flow
    \item $r^+_{gt}$ - Total upward reserve provision
    \item $r^-_{gt}$ - Total downward reserve provision
    \item $r^{L^+}_{glt}$ - Upward reserve provision for reserve level $l$
    \item $r^{L^-}_{glt}$ - Downward reserve provision for reserve level $l$
    \item $rs_{nlt}$ - Upward reserve shedding for reserve level $l$
    \item $rc_{nlt}$ - Downward reserve provided by day ahead load shedding for reserve level $l$
    \item $rinj_{nlt}^{L^+}$ - Possible node injection due to activation of upward reserve level $l$
    \item $rinj_{nlt}^{L^-}$ - Possible node injection due to activation of downward reserve level $l$
    \item $rf_{nbt}^{L^+}$ - Possible branch flow due to activation of upward reserve level $l$
    \item $rf_{nbt}^{L^-}$ - Possible branch flow due to activation of downward reserve level $l$
    \item $d_{nlt}^{L^+}$ - Possible imbalance on node $n$ for upward reserve level $l$
    \item $d_{nlt}^{L^-}$ - Possible imbalance on node $n$ for downward reserve level $l$
\end{itemize}

@@

Note the use of the set $G_n$ to only allow generators at node $n$ to contribute to the power balance (similarly for storage technologies). Another way of describing this would have been through an incidence matrix.

@@

The binary variable $y_{st}$ prevents simultaneous charging and discharging. While such unphysical behaviour is not typical, in this case the constraint was deemed necessary since such behaviour could be exploited to reduce congestion. This was also investigated in the results.

@@

At the time of writing, this convex hull restriction led to model infeasibilities and so it was not implemented.

@@

All data used for this exercise can be found at: \href{https://www.dropbox.com/sh/mdvmc082gwng0tr/AABRyc3fZpxAFycmUfZmh8Csa?dl=0}.

@@

The residual load time series, aggregated across all nodes, for this system is plotted in \todo[inline]{Plot this}.

@@

It should be noted that of the four days selected, only day 285 exhibited load shedding when the load was not scaled but operating reserves were present. This finding highlights how day ahead adequacy may occur for reasons other than insufficient capacity aggregated across the network, i.e. analysis of the aggregated residual load (the load net of renewable generation) may be uninformative.

>>>>

\begin{abstract}
    This report outlines an attempt to integrate a resource adequacy and operational security exercise. The idea is to use a unit commitment schedule which is then used as the input to a more detailed power flow model. An additional goal was to make explicit the trade-off between day ahead adequacy and real time reliability. We used an adapted version of KU Leuven's \ac{DUCPR} model which accounts for network constraints in the reserve activation phase for this purpose. However the \ac{DUCPR} model is not able to capture enough detail in the reserve activation phase for this tradeoff to be made. This and other omissions of the \ac{DUCPR} model when compared to the power flow model meant that the load shedding observed in the two models differed by an order of magntiude. All code for the adequacy side of this exercise can be found at \href{https://github.com/junglegobs/ASoSEPOC}{https://github.com/junglegobs/ASoSEPOC}.
\end{abstract}

@@

Adequacy assessments and operational security analyses are two fundamental exercises in determining and ensuring the reliable operation of the electric power system. The former assesses the ability of the power system to satisfy demand for electricity in the day ahead stage, while the latter assesses (or ensures, depending on the context) that an unexpected event, such as an outage of a transmission line or generator or lower or greater than expected realisation of demand or renewable generation, will lead to cascading failures and a possible blackout.

Typically, these exercises are performed independently. This interaction combines the two by feeding the unit commitment of an adequacy assessment to an operational security analysis which takes these commitments as given.

@@

Secondly, we perform the adequacy assessment using an unconventional \ac{UC} model, a modified version of the \ac{DUCPR} model described in~\cite{Bruninx2017}. This model can be seen as a compromise between a computationally expensive stochastic \ac{UC} model and a more tractable though less accurate deterministic \ac{UC} model with operating reserve requirements. The treatment of operating reserves in this model in principle allows a trade-off between shedding load in the day ahead stage, which would reduce day ahead adequacy, and shedding operating reserves, which should reduce operational security. By making this trade-off explicit and verifying the results in an operational security analysis we wished to elaborate on the results of~\cite{Hermans2018}, which did not consider the possibility that shedding operating reserves would lead to insecure operation of the electric power system.

As noted in the abstract, we were unsuccesful in illustrating the value of this interaction for two reasons. Firstly, we were unable to illustrate the tradeoff between adequacy and security with only the \ac{DUCPR} model due to its approximate treatment of real time network imbalances due to reserve activation. Secondly and related, the expected load shedding resulting from the \ac{DUCPR} differed by an order of magnitude to that found by the power flow, likely due to due to the difference in the treatment of storage as well as the more detailed network representation in the power flow model.

@@

The rest of this document is organised as follows. \secref{sec:model} describes the \ac{DUCPR} model used to simulate power system operations and obtain day ahead adequacy and very approximate real time operational security indicators, namely scheduled load shedding and reserve shedding. \secref{sec:data} describes the stylised Belgian system analysed. \secref{sec:initial_results} analyses a single day in order to identify sources of adequacy and security issues, while \secref{sec:adequacy_vs_security} compares the \ac{DUCPR} model results to that of the power flow analysis.

@@

@@

\begin{itemize}
    \item $G$ - Generators
    \item $G_n$ - Generators located at node $n$
    \item $GD$ - Dispatchable / thermal / conventional generators
    \item $GR$ - Renewable / variable generators (\ac{VRES}).
    \item $S$ - Storage technologies
    \item $S$ - Storage technologies located at node $n$
    \item $N$ - Nodes or buses in the network
    \item $B$ - Lines or branches
    \item $T$ - Time steps / intervals / slices
    \item $L^+$ - Upward reserve levels
    \item $L^-$ - Downward reserve levels
\end{itemize}

@@

\begin{itemize}
    \item $D_{nt}$ - Demand [MW]
    \item $PTDF_{nl}$ - Power Transfer Distribution Factor [-]
    \item $AF_{gt}$ - Availability Factor [-]
    \item $K_g$ - Capacity [MW]
    \item $P^{min}_g$ - Minimum power output [MW]
    \item $P^{max}_g$ - Maximum power output [MW]
    \item $MUT_g$ - Minimum up time [timesteps]
    \item $MDT_g$ - Minimum down time [timesteps]
    \item $RAMP_g$ - Maximum ramping limit (up and down) [MW/timestep]
    \item $D^{L^+}_{lt}$ - Upward reserve requirement [MW]
    \item $D^{L^-}_{lt}$ - Downward reserve requirement [MW]
    \item $P^{L^+}_{lt}$ - Upward reserve level activation probability [-]
    \item $P^{L^-}_{lt}$ - Downward reserve level activation probability [-]
\end{itemize}

@@

\begin{itemize}
    \item $z_{gt}$ - Commitment [-]
    \item $v_{gt}$ - Start up [-]
    \item $w_{gt}$ - Shut down [-]
    \item $q_{gt}$ - Generation [MW]
    \item $\hat{q}_{gt}$ - Generation above the minimum stable operating point (0 in the case of renewables) [MW]
    \item $d_{st}$ - Storage discharge [MW]
    \item $c_{st}$ - Storage charge [MW]
    \item $y_{st}$ - Binary variable, 1 if storage discharges and 0 if it charges [-]
    \item $e_{st}$ - Storage state of charge variable [MWh]
    \item $ls_{nt}$ - Load shedding [MW]
    \item $inj_{nt}$ - Node injection [MW]
    \item $f_{lt}$ - Branch flow [MW]
    \item $r^+_{gt}$ - Total upward reserve provision [MW]
    \item $r^-_{gt}$ - Total downward reserve provision [MW]
    \item $r^{L^+}_{glt}$ - Upward reserve provision for reserve level $l$ [MW]
    \item $r^{L^-}_{glt}$ - Downward reserve provision for reserve level $l$ [MW]
    \item $rs_{nlt}$ - Upward reserve shedding for reserve level $l$ [MW]
    \item $rc_{nlt}$ - Downward reserve provided by day ahead load shedding for reserve level $l$ [MW]
    \item $rinj_{nlt}^{L^+}$ - Possible node injection due to activation of upward reserve level $l$ [MW]
    \item $rinj_{nlt}^{L^-}$ - Possible node injection due to activation of downward reserve level $l$ [MW]
    \item $rf_{nbt}^{L^+}$ - Possible branch flow due to activation of upward reserve level $l$ [MW]
    \item $rf_{nbt}^{L^-}$ - Possible branch flow due to activation of downward reserve level $l$ [MW]
    \item $d_{nlt}^{L^+}$ - Possible imbalance on node $n$ for upward reserve level $l$ [MW]
    \item $d_{nlt}^{L^-}$ - Possible imbalance on node $n$ for downward reserve level $l$ [MW]
\end{itemize}

@@

Note the use of the set $G_n$ to only allow generators at node $n$ to contribute to the power balance (similarly for storage technologies). Another way of describing this would have been through an incidence matrix~\cite{Belderbos2020}.

\subsubsection{Limiting day ahead load shedding}

@@

The binary variable $y_{st}$ prevents simultaneous charging and discharging. While such unphysical behaviour is not typical, in this case the constraint was deemed necessary since such behaviour could be exploited to reduce congestion. This was also investigated in the results, where including this binary variable is referred to as PSCD (`Prevent Simultaneous Charge / Discharge').

@@

This convex hull restriction led to model infeasibilities and so it was not implemented.

@@

All data used for this exercise can be found \href{https://www.dropbox.com/sh/mdvmc082gwng0tr/AABRyc3fZpxAFycmUfZmh8Csa?dl=0}{here}.

@@

The residual load time series, aggregated across all nodes, for this system is plotted in \figref{fig:residual_load_ts}.

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9/textwidth]{../plots/process_input/residual_load_timeseries}
    \caption{Residual load time series of test system.\label{fig:residual_load_ts}}
\end{figure}

@@

Of the four days selected, only day 285 exhibited load shedding when the load was not scaled but operating reserves were present. This finding highlights how day ahead adequacy may occur for reasons other than insufficient capacity aggregated across the network, i.e. analysis of the aggregated residual load (the load net of renewable generation) may be uninformative.

We changed the days under investigation several times during this interaction. For this reason (and also for brevity), we only analysed day 285 in detail (see \secref{sec:initial_results}) and only have a comparison of the \ac{DUCPR} and power flow results for day 309.
